### 2017-03-21 Manuscripts/cog-psych/cogpsych.tex
edits to manuscript

Previous work has focused on one key aspect of category generation: People tend to create new categories that have similar {\em statistical regularities} to previously known categories \citep{jern2013probabilistic,ward1994structured}. Although this is an important characteristic of generating new categories, it cannot be the only one. Taken to the extreme, the best "new" category in terms of having the same statistical regularities to other categories would be one that is identical to a prototypical previously observed category (and thus, not new at all).

In this article, we propose a new constraint that guides category generation: "being different" or contrasting from other categories in the relevant domain. To some extent, new categories being different in some sense from previously observed ones has been implicitly assumed in much previous work. But "being different" is vague as there are many ways to be different from a previously observed category. Building on the largely successful exemplar modeling framework \citep{medin1978context,nosofsky1984choice}, we propose a novel exemplar model of category generation, {\em Producing Alike and Contrasting Knowledge using Exemplar Representations} (PACKER), that formalizes the constraint of "being different" from previous categories while similar to other exemplars generated so far in the new category.

The outline of the article is as follows. First we describe the results of previous behavioral experiments investigating category generation. Next we review previous computational models of category generation. Then we propose our computational model, which produces new categories that are different from previous categories in the domain. Next we present two experiments examining how people generate new categories, one of which demonstrates that participants produce different categories when given one of two categories that do not differ in distributional information. We qualitatively and quantitatively analyze the performance of each model in capturing human category generation. We conclude with a discussion of the implications of our results for categorization and creative cognition, and directions for future work.

@@

Much of the work from this area \citep[e.g.,][]{smith1993constraining,marsh1999inadvertent} focuses on how sample cues (such as an example of a species generated by other participants) can drastically diminish creativity. However, the broader set of observations made by Ward \& colleagues provide a great deal of insight into the nature of creative generation. They indicate that people rely strongly on prior knowledge in the creative process, and people generate concepts in accord with what they already know.

Theoretical accounts of these effects have primarily been grounded within the categorization literature. For example, the predominant ``Path of Least Resistance'' account \citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when generating a new species of animal, people retrieve from memory a known subcategory of animals (e.g., \textit{bird}, \textit{dog}, \textit{horse}), and simply change some of the features to make something new. People were thought to change only features that are not characteristic of the retrieved category (e.g., if \textit{bird} was retrieved, the presence of \textit{wings} would not change, but \textit{color} might). This theory incorporates elements of the highly influential basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as well as the exemplar view \citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. Problematically, however, the experimental paradigms employed in these studies were relatively uncontrolled, precluding the development of formal approaches. Specifically, participants in these studies are typically asked to generate objects that they have a rich degree of prior knowledge about (e.g., plants, animals, tools, toys), and so it is difficult to apply formal models to the data.

@@

The authors evaluated the predictions of several formal models on their data. Most notably, they showed that a Bayesian hierarchical sampling model provided the strongest account. Their model represents categories as multivariate normal distributions within the space, and from exposure to category members the model induces a domain-wide representation of the distributional commonalities among the observed categories. When generating a new category, the model samples a new distribution from the domain, and thus the distributional properties of observed categories will be preserved within the generates category.

\nbcnote{Maybe we need a more technical view of the model here?}

\jlanote{Yeah, Here's how I would this section should be organized. Top is general behavioral findings/phenomena. Then subsections for each theory, which first is described in broad theoretical terms and then how that is formalized into a model}

\cite{jern2013probabilistic} additionally tested a "copy-and-tweak" model that broadly resembles the earlier "Path of Least Resistance" account: the core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice}: participants are thought to generate new items according to their similarity to known members of the target category. This model provided a poor account of the observed data: indeed, the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge this model. However, its application is notable as a first step toward explaining category generation using well-known formal approaches from the categorization literature.

@@

Although it is evident that people are \textit{capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical sampling model developed by \cite{jern2013probabilistic}, for example, does not provide any strong claims about how generated categories should contrast with learned ones: the model only proposes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The authors do not make predictions about the \textit{location} of the category within the domain (the perceptual instantiation of category members). Indeed, under a strict interpretation of the hierarchical Bayesian approach presented by \citeauthor{jern2013probabilistic}, the most probable new category to be generated is located in {\em exactly} the same location as the given category. However, this is not simply an issue with their model -- Hierarchical Bayesian models assume the same underlying distribution generates all of the categories and thus, any differences between categories is due to {\em noise} and should not be {\em systematic}. The best a a hierarchical Bayesian model can do at capturing contrast is to assume that the new category is placed uniformly at random over stimulus space (this was the case in the simulations conducted by \citeauthor{jern2013probabilistic}).

\nbcnote{Maybe tune that language to be more in line with the probabilistic world?}

\jlanote{Don't worry about that for now. I'll do a pass on that sort of language later. Lets get the main stuff in here in the right places first.}

@@

\nbcnote{Maybe mention something about path of least resistance prediction that contrast is on non-characteristic dimensions? }
\jlanote{If we can make a strong case via verbal argument, than yeah. I think we decided not to model it, in which case we shouldn't go into detail on it.}

\section{Formalizing category generation}
\jlanote{this should come before the studies as formalizations of the different theories (some details may be delegated to Appendices). I like trying to integrate them with the verbal description of the theories rather than having a separate computational modeling section. But the latter works for sure and I'm not sure if the former will flow well...}
\subsection{PACKER}
The PACKER model is an extension of the influential Generalized Context Model of category learning \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}. The GCM simulates categorization under the assumption that learners represent categories as a collection of exemplars, corresponding to the labeled stimuli they have observed. The exemplars are encoded within a $k$-dimensional psychological space, and categorization is based on the amount of similarity between the item to be categorized and the stored exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is computed as an inverse exponential function of distance \citep[following][]{shepard1957stimulus,shepard1987toward}:

\begin{equation}
s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k \left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and $\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling the spread of exemplar generalization. For simplicity, attention will be distributed uniformly in our simulations (unless otherwise noted). The value of $r$ depends on the nature of the experimental conditions being simulated: $r=1$ is appropriate for separable dimensions, whereas $r=2$ is appropriate for integral dimensions \citep[see][]{shepard1964attention,garner1974processing}.

@@

\nbcnote{It is worth noting that PACKER is not the only \textit{possible} model that could incorporate a contrast mechanism. However, contrast falls out naturally from the exemplar view and ...  Contrast could, for example, be implemented within... }

\subsection{Relation Between PACKER and Copy-And-Tweak}

@@

\subsection{The Hierarchical Sampling Model}

@@

Research on the creative generation of novel concepts has focused on the finding that generated categories tend to possess distributional commonalities with known categories. However, a fundamental goal of concept generation is to create something \textit{new} (i.e., different from what is already known). The manner in which generated categories differ from known ones is, nonetheless, poorly understood: existing formal approaches do not make strong predictions about how creatively generated concepts should systematically differ from existing ones.

In the sections below, we present two experiments demonstrating systematic effects of category contrast on creative generation. Our experiments conform strictly to the artificial category learning paradigm from \cite{jern2013probabilistic}: participants are first exposed to a single, experimenter-defined category, and are then asked to generate members of a new category. We also introduce a novel, exemplar-based model to explain these effects: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). Finally, we report formal simulations comparing PACKER to the hierarchical sampling and copy-and-tweak models developed by \cite{jern2013probabilistic}.

@@

To begin our investigation, we developed a artificial, two dimensional domain of squares, varying in color and size (see Figure \ref{fig:sample-stimuli} for samples). We created three unique category types within this domain, shown in Figure \ref{fig:e1-conditions}. The category types were designed to possess strong differences in the distributional structure of the experimenter-defined (`A') category, allowing us to observe the effect of category contrast under a variety of types of prior knowledge. These conditions also provide an opportunity to replicate classic effects observed in previous reports \citep[e.g.,][]{jern2013probabilistic,ward1994structured,ward2002role,marsh1999inadvertent,smith1993constraining}.

\nbcnote{This feels a little weak in motivation: how do these conditions specifically allow us to show is these effects? Did we come up with these out of thin air? So maybe foreshadow that you hardly need to look for these effects, they're \textit{everywhere}. Maybe there is also a problem that at this point it is not exactly clear what contrast means within this scenario.}

\jlanote{Is it possible to get the main individual differences we find in this study from different parameterizations of PACKER? If so, we can motivate it from there. Alternatively, we may want to switch the order of the two studies and motivate the experiment from the next one as a . I like switching the order of the experiments less because I feel like Experiment 2 nails contrast down and Experiment 1 is more showing how it plays a role in more complex category types}

@@

In the `Cluster' type, the experimenter-defined category is a tight cluster of examples in the space (e.g., all members of category `A' are small and dark in color). In the `Row' type, the category has a row pattern across the space: varying along one feature but not the other (e.g., members of category `A' are small and vary in color). In the `XOR' type, the experiment-defined category consists of two clusters separated in opposite corners of the space category, conforming to the exclusive-or logical structure (e.g., members of category `A' can be small and dark or large and light).

@@

>>>>

Previous work has focused on one key aspect of category generation: People tend to create new categories that have similar {\em statistical regularities} to previously known categories \citep{jern2013probabilistic,ward1994structured}. Although this is an important characteristic of generating new categories, it cannot be the only one. Taken to the extreme, the best ``new'' category in terms of having the same statistical regularities to other categories would be one that is identical to a prototypical previously observed category (and thus, not new at all).

In this article, we propose a new constraint that guides category generation: ``being different'' or contrasting from other categories in the relevant domain. To some extent, this constraint has been implicitly assumed in much previous work: the idea that a new category should be ``different'' is vague, as there are many ways it could be different from a previously observed category. Building on the largely successful exemplar modeling framework \citep{medin1978context,nosofsky1984choice,nosofsky1986attention}, we propose a novel exemplar model of category generation, {\em Producing Alike and Contrasting Knowledge using Exemplar Representations} (PACKER), formalizing how new categories should differ from previous categories.

The outline of the article is as follows. First we describe previous empirical work on the topic of category generation, as well as the computational approaches studied in those reports. Then we describe our novel computational model, which is designed to generate categories that systematically differ from existing categories in the domain. We present two experiments demonstrating strong and systematic effects of category contrast on creative generation, and we qualitatively and quantitatively analyze the performance of each model in capturing human category generation. We conclude with a discussion of the implications of our results for categorization and creative cognition, and directions for future work.

@@

Much of the work from this area \citep[e.g.,][]{smith1993constraining,marsh1999inadvertent} focuses on how sample cues (such as an example of a species generated by other participants) can drastically diminish creativity. However, the broader set of observations made by Ward \& colleagues provide a great deal of insight into the nature of creative generation. They indicate that people rely strongly on prior knowledge in the creative process, and people generate concepts in accord with what they already know. 
,
Theoretical accounts of these effects have primarily been grounded within the categorization literature. For example, the predominant ``Path of Least Resistance'' account \citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when generating a new species of animal, people retrieve from memory a known subcategory of animals (e.g., \textit{bird}, \textit{dog}, \textit{horse}), and simply change some of the features to make something new. People were thought to change only features that are not characteristic of the retrieved category (e.g., if \textit{bird} was retrieved, the presence of \textit{wings} would not change, but \textit{color} might). This theory incorporates elements of the highly influential basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as well as the exemplar view \citep{medin1978context,brooks1978nonanalytic}. Problematically, however, the experimental paradigms employed in these studies were relatively uncontrolled, precluding the development of formal approaches. Specifically, participants in these studies are typically asked to generate objects that they have a rich degree of prior knowledge about (e.g., plants, animals, tools, toys), and so it is difficult to apply formal models to the data.

@@

The authors evaluated the predictions of several formal models on their data. Most notably, they showed that a Bayesian hierarchical sampling model provided the strongest account. Their model views observed examples as samples from an underlying category distribution, describing the location of the category in the space, as well as how it varies along each feature. In turn, each category is viewed as a sample from an underlying \textit{domain} distribution, specifying distributional commonalities among the observed categories. Generated categories are thought to stem from the same domain distribution as observed categories, thus the distributional properties of observed categories will be preserved within the generated category.

% Here's what it used to be:
% Their model represents categories as multivariate normal distributions within the space, and from exposure to category members the model induces a domain-wide representation of the distributional commonalities among the observed categories. When generating a new category, the model samples a new distribution from the domain, and thus the distributional properties of observed categories will be preserved within the generates category.

% \jlanote{Yeah, Here's how I would this section should be organized. Top is general behavioral findings/phenomena. Then subsections for each theory, which first is described in broad theoretical terms and then how that is formalized into a model}

\cite{jern2013probabilistic} additionally tested a "copy-and-tweak" model that broadly resembles the earlier "Path of Least Resistance" account: the core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}: participants are thought to generate new items according to their similarity to known members of the target category. This model provided a poor account of the observed data: indeed, the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge this model. However, its application is notable as a first step toward explaining category generation using well-known formal approaches from the categorization literature.

@@

% Joe says he'll do a pass on this language later:
Although it is evident that people are \textit{capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical sampling model developed by \cite{jern2013probabilistic}, for example, does not provide any strong claims about how generated categories should contrast with learned ones: the model only proposes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The authors do not make predictions about the \textit{location} of the category within the domain (the perceptual instantiation of category members). Indeed, under a strict interpretation of the hierarchical Bayesian approach presented by \citeauthor{jern2013probabilistic}, the most probable new category to be generated is located in {\em exactly} the same location as the given category. However, this is not simply an issue with their model -- Hierarchical Bayesian models assume the same underlying distribution generates all of the categories and thus, any differences between categories is due to {\em noise} and should not be {\em systematic}. The best a a hierarchical Bayesian model can do at capturing contrast is to assume that the new category is placed uniformly at random over stimulus space (this was the case in the simulations conducted by \citeauthor{jern2013probabilistic}).

@@

% \section{Formalizing category generation}
% \jlanote{this should come before the studies as formalizations of the different theories (some details may be delegated to Appendices). I like trying to integrate them with the verbal description of the theories rather than having a separate computational modeling section. But the latter works for sure and I'm not sure if the former will flow well...}

\subsection{The PACKER Model}

As noted above, the constraint that new concepts should differ from what is already known has been largely been overlooked in previous work. This is no doubt in part due to the vague definition of what it means for a concept to be ``different'': a generated category may be different from what is already known in any number of respects. Towards formalizing the role of contrast in creative generation, we developed the PACKER  ({\em Producing Alike and Contrasting Knowledge using Exemplar Representations}) model which explains category generation as a balance between two fundamental constraints: generated categories should not be similar to known categories, and exemplars within each category should be similar to one another. These ideas are implemented within the well-studied exemplar framework: the PACKER model is an extension of the influential Generalized Context Model of category learning \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

\nbcnote{Maybe note something about how PACKER's basis in the exemplar view means that we're explaining category generation using well known ideas, and that exemplar models are viewed as highly principled.}

Both PACKER and the GCM simulate categorization under the assumption that learners represent categories as a collection of exemplars, corresponding to the labeled stimuli they have observed. The exemplars are encoded within a $k$-dimensional psychological space, and model performance is based on the amount of similarity between the item to be categorized and the stored exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is computed as an inverse exponential function of distance \citep[following][]{shepard1957stimulus,shepard1987toward}:

\begin{equation}
s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k \left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and $\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling the spread of exemplar generalization. For simplicity, attention will be distributed uniformly in our simulations (unless otherwise noted). The value of $r$ depends on the nature of the experimental conditions being simulated: $r=1$ is appropriate for separable dimensions, whereas $r=2$ is appropriate for integral dimensions \citep[see][]{shepard1964attention,garner1974processing}. In our simulations, we set $r=1$ due to the separable nature of the stimulus dimensions used in our experiments (see Figure \ref{fig:sample-stimuli}).

@@

It is worth noting that PACKER represents just one of many possible models that incorporate contrast in category generation. For example, although it may be possible to implement a contrast mechanism within \cite{jern2013probabilistic}'s hierarchical sampling model, as noted above this mechanism would be at odds with a strict interpretation of the hierarchical Bayesian framework. In contrast, these ideas emerge naturally from the exemplar view: we have not modified any of the core elements of the GCM in defining the PACKER model, we simply aggregated similarity slightly differently. \nbcnote{That line about importance sampling can go here.}

\subsubsection{Relation Between PACKER and Copy-And-Tweak}

@@

@@

Research on the creative generation of novel concepts has focused on the finding that generated categories tend to possess distributional commonalities with known categories. However, a fundamental goal of concept generation is to create something \textit{new} (i.e., different from what is already known). The manner in which generated categories differ from known ones is, nonetheless, poorly understood: existing formal approaches do not make strong predictions about how creatively generated concepts should systematically differ from existing ones. Above, we introduced a novel, exemplar-based model formalizing the role of contrast in creative generation.

In the sections below, we present two experiments demonstrating systematic effects of category contrast on creative generation. Our experiments conform strictly to the artificial category learning paradigm from \cite{jern2013probabilistic}: participants are first exposed to a single, experimenter-defined category, and are then asked to generate members of a new category. We then we report formal simulations comparing PACKER's account of our results to that of the hierarchical sampling and copy-and-tweak models developed by \cite{jern2013probabilistic}.

@@

To begin our investigation, we developed a artificial, two dimensional domain of squares, varying in color and size (see Figure \ref{fig:sample-stimuli} for samples). To foreshadow slightly, effects of contrast are strong and can be widely observed in category generation data. Accordingly, we began by testing the strength of these effects under a variety of learning conditions. This was achieved by training participants on categories possessing qualitatively different distributional structures: We created three unique category types within the domain, shown in Figure \ref{fig:e1-conditions}.

Each panel of Figure \ref{fig:e1-conditions} shows the locations of exemplar belonging to a single, experimenter-defined category (`A', or `Alpha'). In the `Cluster' type, category A is a tight cluster of examples in the space (e.g., all members are small and dark in color). In the `Row' type, the category has a row pattern across the space: varying along one feature but not the other (e.g., members are small and vary in color). In the `XOR' type, the experiment-defined category consists of two clusters separated in opposite corners of the space category, conforming to the exclusive-or logical structure (e.g., members can be small and dark or large and light).

\nbcnote{what would contrast look like?}

% \nbcnote{This feels a little weak in motivation: how do these conditions specifically allow us to show is these effects? Did we come up with these out of thin air? So maybe foreshadow that you hardly need to look for these effects, they're \textit{everywhere}. Maybe there is also a problem that at this point it is not exactly clear what contrast means within this scenario.}
% \jlanote{Is it possible to get the main individual differences we find in this study from different parameterizations of PACKER? If so, we can motivate it from there. Alternatively, we may want to switch the order of the two studies and motivate the experiment from the next one as a . I like switching the order of the experiments less because I feel like Experiment 2 nails contrast down and Experiment 1 is more showing how it plays a role in more complex category types}

@@

@@

\appendix

\section{The Hierarchical Sampling Model}
\label{ap:hsampling-definition}

\nbcnote{It \textbf{is} just conjugacy, so we can just cite a paper (do you know of a paper or should I start digging?). We'll probably want to detail the what we did for priors and which free parameters we fitted though.}
